---
title: "Spotify daily top 200 songs"
date: "10 11 2021"
output: 
  prettydoc::html_pretty:
    theme: cayman
    toc: true
  
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(plotly)
library(dplyr)
library(lubridate)
library(stringr)
library(data.table)
library(RColorBrewer)
library(wordcloud)
library(kableExtra)

spotify <- read.csv2("~/Documents/OneDrive/Musik/SpotifyTop200/SpotifyDailyTop200-2017_21.csv", sep="#")
#spotify <- read.csv2("~/OneDrive/Musik/SpotifyDailyTop200-2017_21.csv", sep="#")

```

On this chapter I will explore the following dataset available on kaggle: 

- [Spotify daily top 200 songs with genres 2017-2021](https://www.kaggle.com/ivannatarov/spotify-daily-top-200-songs-with-genres-20172021)

It contains information on the daily top 200 tracks listened to by users of the Spotify digital platform around the world.

Here a quick glance at the data as is:

```{r echo=FALSE}
kable(head(spotify)) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F)
```
**Note:** For this exercise 2021 will not be included, as the year had not ended when closing this page.  
  
  
## First impressions
  
The amount of streams has been increasing over the years as willingnes to pay for the service also increases and anti-piracy measures are being enforced in most countries.
  
We see more artists reaching the top 200, always with many cool songs to move us.

```{r echo=FALSE, fig.show='hold', out.width='33%'}

spotify<-mutate(spotify,Year=year(spotify$Date))
spotify<-mutate(spotify,Month=month(spotify$Date))
spotify<-subset(spotify, Year < 2021)

aggregate(spotify$Streams, by=list(spotify$Year), FUN=sum)%>%
  mutate(x = x/1000000000) %>%
  ggplot(aes(factor(Group.1),x,alpha=0.3, fill=factor(Group.1)))+
  geom_col()+
  geom_text(aes(label=round(x, 1)), size= 8, position = position_stack(vjust = 0.5))+
  labs(title = "Billion of Streams each year...", y=NULL, x=NULL)+
  theme_light()+
  theme(text = element_text(size = 20),legend.position = "none", axis.ticks.x = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  scale_fill_manual(values = c("#25bf1d","#25bf1d","#25bf1d","#25bf1d","#25bf1d"), aesthetics = "fill")

spotify %>%
  group_by(Year) %>%
  summarise(ArtistCount = n_distinct(Artist)) %>%
  ggplot(aes(factor(Year), ArtistCount, alpha=0.3, fill=factor(Year))) +
  geom_col()+
  geom_text(aes(label=ArtistCount), size= 8, position = position_stack(vjust = 0.5))+
  labs(title = "... of talented Artists...", y=NULL, x=NULL)+
  theme_light()+
  theme(text = element_text(size = 20), legend.position = "none", axis.ticks.x = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  scale_fill_manual(values = c("#25bf1d","#25bf1d","#25bf1d","#25bf1d","#25bf1d"), aesthetics = "fill")

spotify %>%
  group_by(Year) %>%
  summarise(TrackCount = n_distinct(Track.Name))%>%
  ggplot(aes(factor(Year), TrackCount, alpha=0.3, fill=factor(Year))) +
  geom_col()+
  geom_text(aes(label=TrackCount), size= 8, position = position_stack(vjust = 0.5))+
  labs(title = "... with inspiring songs", y=NULL, x=NULL)+
  theme_light()+
  theme(text = element_text(size = 20), legend.position = "none", axis.ticks.x = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  scale_fill_manual(values = c("#25bf1d","#25bf1d","#25bf1d","#25bf1d","#25bf1d"), aesthetics = "fill")
```

These artists made it to the top 3 of each year  
  
```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='100%'}
top3 <- spotify %>%
  group_by(Year, Artist) %>%
  summarise(TotalStreamsYr = sum(Streams)) %>%
  arrange(Year, desc(TotalStreamsYr)) %>%
  top_n(3) %>%
  left_join(spotify %>%
              group_by(Year, Artist, Track.Name) %>%
              summarise(TrackStreamsYr = sum(Streams))
            , by=c("Year", "Artist")) %>%
  arrange(Year, desc(TotalStreamsYr), desc(TrackStreamsYr)) %>%
  mutate(TotalStreamsYr=round(TotalStreamsYr/1000000000,1), TrackStreamsYr=round(TrackStreamsYr/1000000000,1))

top3 <- setDT(top3)[ , .SD[which.max(TrackStreamsYr)], by = c("Year", "Artist")]
  
ggplot(top3, aes(reorder(Artist, -TotalStreamsYr), TotalStreamsYr, fill=factor(Year), group=Year)) + 
  geom_bar(stat = "Identity", position = "stack", alpha=0.35) +
  geom_text(aes(label=TotalStreamsYr), size=2, angle=45)+
  geom_col(aes(Artist,TrackStreamsYr, fill=factor(Year)))+
  geom_text(aes(label=paste(Track.Name,TrackStreamsYr, sep = "\n")), position = position_stack(vjust = 0.1), size=2, angle=45)+
  theme_light()+
  theme(text = element_text(size = 8),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(),axis.text.x=element_text(angle=90, hjust=1))+
  labs(title = "Top 3 Artists", y="Streams (bn)", x=NULL)+
  facet_grid(~Year, scales = "free")
```

## Genres in a cloud

```{r echo=FALSE, warning=FALSE, message=FALSE}
Genres <- spotify #[spotify$Position<=10,]
Genres <- aggregate(x=Genres$Streams, by=list(Genres$Year, Genres$Genre), FUN=sum)
colnames(Genres) <- c("Year","Genre","Streams")
Genres$Genre <- str_sub(Genres$Genre,2, -2)
Genres$Genre <- str_remove_all(Genres$Genre,"'")
Genres$Genre <- str_replace_all(Genres$Genre, ", ", ",")
TopGenre <- data.frame(str_split(Genres$Genre, ",", simplify = TRUE))
TopGenre <- data.frame(bind_cols(TopGenre, Genres))
TopGenre$Genre <-  NULL
TopGenre <- tidyr::gather(TopGenre,key="OriginalCol",value="Genre",-Year, -Streams)
TopGenre$OriginalCol <- NULL
TopGenre <-  aggregate(x=TopGenre$Streams, by=list(TopGenre$Year, TopGenre$Genre), FUN=sum)
colnames(TopGenre) <-  c("Year","Genre","Streams")
TopGenre <- TopGenre[TopGenre$Genre != "",]
rm(Genres)
TopGenreTotal <- aggregate(x=TopGenre$Streams, by=list(TopGenre$Genre), FUN=sum)

wordcloud(words = TopGenreTotal$Group.1, freq = TopGenreTotal$x, max.words = 100,colors=brewer.pal(12, "Paired"))
```
  
It's hard to see the forest for the trees, so let's pick the top 3 Genres for each year
  
```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='70%'}
TopGenre %>%
  arrange(Year, desc(Streams)) %>%
  top_n(12) %>%
  mutate(Streams=round(Streams/1000000000,1)) %>%
  ggplot(aes(Year, Streams, fill=Genre)) +
  geom_col(alpha=0.5)+
  geom_text(aes(label=paste(Genre,Streams, sep = "\n")), position = position_stack(vjust = 0.5), size=4)+
  theme_light()+
  theme(text = element_text(size = 12),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  labs(title = "Top 3 Genres per year", y="Streams (bn)", x=NULL, fill="Top genre")
```

## What else?
### Extending the dataset
#### Scraping the web
I realised in the dataset there was no indication to the duration of the track, so I decided to write a little webscraping code to get this info from the [Last.fm](www.last.fm) database.
  
  Here the code I wrote if you're into techy stuff. If not, just scroll down =P
  
```{r echo=TRUE, eval=FALSE}
# Import the necessary libraries 
library(rvest)
library(dplyr)
library(stringr)

# Get the Track and Artists names
DurationList <- unique(spotify[c("Track.Name","Artist")])
DurationList$Duration <-  0

# Replace some characters
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name, " ", "+")
DurationList$Artist <- str_replace_all(DurationList$Artist, " ", "+")
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name,"'","%27")
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name,"\"","%28")

# Run over the Last.fm search pages and retrieve the Duration of each Track
for (i in 1:5149){
  tryCatch({
    link = paste0("https://www.last.fm/search?q=", DurationList[i,1], "+", DurationList[i,2])
    page = read_html(link)
    DurationList[i,3] <- page %>% html_nodes(".chartlist-duration") %>% .[1] %>% html_text() %>%
      str_replace_all(" ", "") %>% str_replace_all("\\n", "")
    print(paste0(i, " ",DurationList[i,2], " ",DurationList[i,1]," ", DurationList[i,3]))
  }, error = function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# Get the duration in seconds as the mm:ss format wouldn't allow for calculations
DurationList$DurationSec <- str_split(DurationList$Duration, pattern = ":", simplify = TRUE)
DurationList$DurationSec <- as.numeric(DurationList$DurationSec[,1])*60 + as.numeric(DurationList$DurationSec[,2])

# Now revert the character replacements
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name, "\\+", " ")
DurationList$Artist <- str_replace_all(DurationList$Artist, "\\+", " ")
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name,"%27","'")
DurationList$Track.Name <- str_replace_all(DurationList$Track.Name,"%28","\"")

# Then expand the original dataset by joining with the newly harvested data
tempos <- left_join(spotify,lista, by=c("Track.Name", "Artist"))
# Remove some of the NAs from the scraping
tempos <- tempos[!is.na(tempos$DurationSec),]
```
and here a glance at the final result
```{r echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
lista <- read.csv2("lista3.csv")
library(stringr)
lista$DurationSec <- str_split(lista$Duration, pattern = ":", simplify = TRUE)
lista$DurationSec <- as.numeric(lista$DurationSec[,1])*60 + as.numeric(lista$DurationSec[,2])

lista$Track.Name <- str_replace_all(lista$Track.Name, "\\+", " ")
lista$Artist <- str_replace_all(lista$Artist, "\\+", " ")
lista$Track.Name <- str_replace_all(lista$Track.Name,"%27","'")
lista$Track.Name <- str_replace_all(lista$Track.Name,"%28","\"")

tempos <- left_join(spotify,lista, by=c("Track.Name", "Artist"))
tempos <- tempos[!is.na(tempos$DurationSec),]
tempos$X <- NULL
kable(head(tempos)) %>%
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F)
```

### Songs that won't last forever
So now let's check the newly acquired values and how they relate to the other variables.

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='70%'}
tempos %>%
  group_by(Artist, Track.Name, DurationSec) %>%
  summarise(totalStreamsYr = sum(Streams)) %>%
  mutate(totalStreamsYr=round(totalStreamsYr/1000000,1)) %>%
  ggplot(aes(DurationSec,totalStreamsYr)) +
  geom_point(aes(alpha=0.01))+
  theme(legend.position = "none")+
  theme_light()+
  labs(title = "Streams per Song", subtitle = "Song legth", y="Total Streams (M)", x="Length (s)", caption = "Each dot represents one singular song.")+
  theme(text = element_text(size = 10),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.caption.position =  "plot", plot.caption = element_text(hjust = 0, face= "italic"))
```

That looks quite cluttered, so let's build a quick histogram to see how many times a given track length occurs in the new dataset.

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='70%'}
tempos %>%
  ggplot(aes(DurationSec))+
  geom_histogram(aes(y =..density.., alpha=1), binwidth = 20)+
  stat_function(fun = dnorm, args = list(mean = mean(tempos$DurationSec), sd = sd(tempos$DurationSec)), col="red")+
  theme(legend.position = "none")+
  theme_light()+
  theme(text = element_text(size = 10),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  labs(title = "Histogram with normal density curve", y="Density", x="Length (s)")
```

It may be interesting to break down the density per year.

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='70%'}
tempos %>%
  group_by(Year, Artist, Track.Name, DurationSec) %>%
  summarise(totalStreamsYr = sum(Streams)) %>%
  mutate(totalStreamsYr=round(totalStreamsYr/1000,1)) %>%
  ggplot(aes(x=DurationSec, colour=as.factor(Year), fill=as.factor(Year))) +
  geom_density(alpha=0.03)+
  xlim(0,400)+
  theme_light()+
  theme(text = element_text(size = 10),legend.position = "bottom", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  labs(title = "Track length distribution", y="Density", x="Length (s)", colour="Year", fill="Year")
```

Looks like the average track length has been decreasing in the year on year comparison, the following boxplots may make it clearer:

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='70%'}
tempos %>%
  filter(DurationSec<400) %>%
  group_by(Year, Artist, Track.Name, DurationSec) %>%
  summarise(totalStreamsYr = sum(Streams)) %>%
  mutate(totalStreamsYr=round(totalStreamsYr/1000000,1)) %>%
  ggplot(aes(totalStreamsYr, DurationSec, colour=as.factor(Year))) +
  geom_boxplot(notch = TRUE) +
  facet_grid(~Year, scales = "free")+
  theme_light()+
  theme(text = element_text(size = 10),legend.position = "bottom", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank())+
  labs(title = "Track length distribution", y="Length (s)", x="Streams (M)", color="Year")
```

An here the explicit year on year decrease in the average length of the tracks:

```{r echo=FALSE, warning=FALSE, message=FALSE}
tempos %>%
  group_by(Year, Artist, Track.Name, DurationSec) %>%
  summarise(totalStreamsYr = sum(Streams)) %>%
  mutate(totalStreamsYr=round(totalStreamsYr/1000000,1)) %>%
  group_by(Year) %>%
  summarise(Average_Length=round(mean(DurationSec), digits = 1)) %>%
  mutate(YoY = round((Average_Length-lag(Average_Length))/Average_Length*100, digits = 1)) %>%
  kable() %>%
  kable_styling(bootstrap_options = "striped", font_size = 10, full_width = F, position = "left")
```

So all in all, we have more songs in the daily Top 200 streamed songs on Spotify, however the average duration has been shrinking over the past four years.

### What are the odds?
Only `r paste(round((tempos%>%
  filter(Position==1)%>%
  distinct(Track.Name, Artist, Genre, DurationSec) %>%
  nrow()
/
tempos%>%
  distinct(Track.Name, Artist, Genre, DurationSec) %>%
  nrow()
)*100, digits=2),"%")` of the songs make it to the very top, and `r paste(round((tempos%>%
  filter(Position<=10)%>%
  distinct(Track.Name, Artist, Genre, DurationSec) %>%
  nrow()
/
tempos%>%
  distinct(Track.Name, Artist, Genre, DurationSec) %>%
  nrow()
)*100, digits=2),"%")` enter the top 10. But does it relate in any way with the length of the track? Looking at the charts below, seems plausible to assume that.

```{r echo=FALSE, warning=FALSE, message=FALSE, out.width='50%'}
tempos %>%
  filter(DurationSec <= 500) %>%
  group_by(DurationSec) %>%
  summarise(HighestPosition = min(Position)) %>%
  ggplot(aes(DurationSec, HighestPosition)) +
  geom_point(aes(alpha=0.5), position = "jitter") +
  geom_density2d_filled(aes(alpha=0.5)) +
  geom_density_2d(size = 0.25, colour = "black") +
  scale_y_reverse(n.breaks=10) +
  theme(legend.position = "none")+
  theme_light()+
  labs(title = "Heatmap", subtitle = "Length vs Ranking Position", y="Ranking (1-200)", x="Length (s)")+
  theme(text = element_text(size = 15),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.caption.position =  "plot", plot.caption = element_text(hjust = 0, face= "italic"))

tempos %>%
  group_by(DurationSec) %>%
  summarise(HighestPosition = min(Position)) %>%
  filter(HighestPosition <= 10) %>%
  ggplot(aes(DurationSec, HighestPosition)) +
  geom_point(aes(alpha=0.5), position = "jitter") +
  geom_density2d_filled(aes(alpha=0.5)) +
  geom_density_2d(size = 0.25, colour = "black") +
  scale_y_reverse(n.breaks=10) +
  theme(legend.position = "none")+
  theme_light()+
  labs(title = "Heatmap (Top 10)", subtitle = "Length vs Ranking Position", y="Ranking (1-10)", x="Length (s)")+
  theme(text = element_text(size = 15),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.caption.position =  "plot", plot.caption = element_text(hjust = 0, face= "italic"))
```

I went on and calculated the probabilities of a given track length to join the top 10 for at least one day then plotted that into the chart below:

```{r echo=FALSE, warning=FALSE, message=FALSE}
Probs <- tempos %>%
  group_by(DurationSec) %>%
  summarise(HighestPosition = min(Position), Count= n()) %>%
  filter(HighestPosition<=10)%>%
  mutate(Probability = (Count / 254071)*100) %>%
  ggplot(aes(DurationSec, Probability)) +
  geom_point(aes(alpha=20)) +
  geom_density_2d(size = 0.25, colour = "grey", alpha=0.5) +
  theme_light() +
  labs(title = "Probability distribution", subtitle = "Tracks that got into the top 10", y="Probability", x="Length (s)") +
  theme(text = element_text(size = 10),legend.position = "none", axis.ticks.x = element_blank(), axis.ticks.y = element_blank(), panel.grid = element_blank(), plot.caption.position =  "plot", plot.caption = element_text(hjust = 0, face= "italic"))
```

Hope you enjoyed this overview. The code used for building this report can be found on my [GitHub](https://github.com/AllysonAOdS/SpotifyTop200).